The impact of gas bulk rotation on the Lyman $\alpha$ line

### 2014-04-15 paper/ms.tex
advances

\title{The impact of gas bulk rotation on the Lyman $\alpha$ line} 
\shorttitle{Rotation effects on the Lyman $\alpha$ line}
\shortauthors{Garavito-Camargo, Forero-Romero \& Dijstra}
\author{ Juan N. Garavito-Camargo, Jaime E. Forero-Romero}
\affil{Departamento de F\'{i}sica, Universidad de los Andes, Cra. 1
No. 18A-10, Edificio Ip, Bogot\'a, Colombia}
\email{jn.garavito57@uniandes.edu.co}
\email{je.forero@uniandes.edu.co}
\and
\author{Mark Dijkstra}
\affil{Institute of Theoretical Astrophysics, University of Oslo, Postboks 1029, 0858 Oslo, Norway}
\email{mark.dijkstra@astro.uio.no }

@@

We present results of radiative transfer calculations to measure the
impact of gas bulk rotation on the morphology of the Lyman $\alpha$
emission line in distant galaxies. We model a galaxy as a sphere with
a homogeneous mixture of dust and hydrogen at a constant
temperature. These spheres have a solid-body rotation with maximum
velocities in the range $0-300$ \kms and neutral hydrogen optical
depths in the range $\tau_{\rm H}=10^{5}-10^{7}$. We consider cases 
of a single central \lya source, and a uniform distribution of sources. 
Our main finding are that the escape fraction of \lya photons, and
the intensity at the line center increases with rotational velocity. 
For homogeneously distributed sources the line becomes single peaked 
at rotational velocities larger than half the line width in the static case.
For radiation sources located off-center, the line morphology presents 
a range of single, double and triple peaks. We show how these results 
are useful to interpret recent spectroscopic results of distant 
$z\sim 2-3$ star forming galaxies.

@@

A quantitative description of the \ly line has been reached through
Monte Carlo simulations \citep{Auer68,Avery68,Adams72}. In the last
two decades these studies have become popular due to the
availability of computing power. Early into the 21st century, the first
studies focused on homogeneous and static media
\citep{Ahn00,Ahn01,Zheng02}; Later on, the effects of clumpy media
\citep{Hansen06} and expanding/contracting shell/spherical geometries started to
be studied \citep{Verhamme06,Dijkstra06,Ahn2014}. Similar codes have applied
these results to semi-analytic models of galaxy formation \citep{Orsi12, GAREL2012} and
results of large hydrodynamic simulations
\citep{CLARA,Forero12,Behrens13}. Recently, Monte Carlo codes have also
been applied to the results of high resolution hydrodynamic
simulations of individual
galaxies\citep{Laursen09,Barnes11,Verhamme12,Yajima12}. Meanwhile, recent
developments have been focused on the systematic study of clumpy
outflows \citep{DijkstraKramer}and anisotropic velocity configurations
\citep{Zheng2013}.

The recent studies of galaxies in hydrodynamic simulations
\citep{Laursen09,Barnes11,Verhamme12,Yajima12} have all shown
systematic variations in the \ly line with the viewing angle. These
variations are a complex superposition of anisotropic density
configurations (i.e. edge-on vs. face-on view of a galaxy), the
inflows observed by gas cooling and the outflows included in the
supernova feedback process of the simulation. These bulk flows
physically correspond to the circumgalactic and intergalactic medium
(CGM and IGM). These effects are starting to be studied
 in simplified configurations that vary the density and wind
 characteristics \citep{Zheng2013,BEHRENS2014}.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.8\textwidth]{f1.eps}
\end{center}
\caption{Shape of the \ly line for
    different maximum rotational velocities. The continuous (dashed)
    line represents the central (homogeneous) source distributions. 
    This profiles were obtained integrating the flux of photons 
	over all directions, this means we made histograms of the 
	numbers of photons escaping the galaxy with a given frequency.
    \label{fig:differentvelocities}}  
\end{figure*}

@@

Are these features observable with current instrumental capabilities?
As an order of magnitude estimate we can consider that the width of
each one of the triple peaks is close to $1/3$ of the FWHM of the line
at a given $V_{\rm max}$. For the cases we have studied in this
subsection those values would be close to $50\kms$, $100\kms$ and
$200\kms$ for optical depths of $10^5$, $10^6$, and $10^7$
respectively. At a redshift of $z\sim 2-3$ each one of these widths
is multiplied by factors of $z+1\sim 3-4$. Considering a nominal value
for the resolution of $\Delta v=200\kms$ for a Near-IR spectrograph capable of observing
those galaxies \citep{Kulas12}, we find that only the case of triple
peaked emission for the most massive object $\tau_{\rm H}, 10^{7}$,
$V_{\rm max}=200\kms$ would have a chance to be spectrally resolved.

@@

@@

\item The spatial anisotropy induced by rotation in the emission is
  usually below the $10\%$ level.

>>>>

\title{The impact of gas bulk rotation on the Lyman $\alpha$ line} 
\shorttitle{Rotation effects on the Lyman $\alpha$ line}

\shortauthors{Garavito-Camargo, Forero-Romero \& Dijstra}

\author{ Juan N. Garavito-Camargo, Jaime E. Forero-Romero}
\affil{Departamento de F\'{i}sica, Universidad de los Andes, Cra. 1
No. 18A-10, Edificio Ip, Bogot\'a, Colombia}
\email{jn.garavito57@uniandes.edu.co}
\email{je.forero@uniandes.edu.co}
\and
\author{Mark Dijkstra}
\affil{Institute of Theoretical Astrophysics, University of Oslo, Postboks 1029, 0858 Oslo, Norway}
\email{mark.dijkstra@astro.uio.no }

@@

We present results of radiative transfer calculations to measure the
impact of gas bulk rotation on the morphology of the Lyman $\alpha$
emission line in distant galaxies. We model a galaxy as a sphere with
a homogeneous mixture of dust and hydrogen at a constant
temperature. These spheres have a solid-body rotation with maximum
velocities in the range $0-300$ \kms and neutral hydrogen optical
depths in the range $\tau_{\rm H}=10^{5}-10^{7}$. We consider cases 
of a single central \lya source, and a uniform distribution of sources. 
Our main finding are that the escape fraction of \lya photons, and
the intensity at the line center increases with rotational velocity. 
For homogeneously distributed sources the line becomes single peaked 
at rotational velocities larger than half the line width in the static
case. We also find that this anisotripic velocity configuration does
not induce any significant anisotropy in the spectral properties.
For radiation sources located off-center, the line morphology
presents  a range of single, double and triple peaks. We show how
these results  are useful to interpret recent spectroscopic results of distant 
$z\sim 2-3$ star forming galaxies.

@@

A quantitative description of the \ly line has been reached through
Monte Carlo simulations \citep{Auer68,Avery68,Adams72}. In the last
two decades these studies have become popular due to the
availability of computing power. Early into the 21st century, the first
studies focused on homogeneous and static media
\citep{Ahn00,Ahn01,Zheng02}; Later on, the effects of clumpy media
\citep{Hansen06} and expanding/contracting shell/spherical geometries started to
be studied \citep{Verhamme06,Dijkstra06,Ahn2014}. Similar codes have applied
these results to semi-analytic models of galaxy formation \citep{Orsi12, Garel2012} and
results of large hydrodynamic simulations
\citep{CLARA,Forero12,Behrens13}. Recently, Monte Carlo codes have also
been applied to the results of high resolution hydrodynamic
simulations of individual
galaxies\citep{Laursen09,Barnes11,Verhamme12,Yajima12}. Meanwhile, recent
developments have been focused on the systematic study of clumpy
outflows \citep{DijkstraKramer}and anisotropic velocity configurations
\citep{Zheng2013}.

The recent studies of galaxies in hydrodynamic simulations
\citep{Laursen09,Barnes11,Verhamme12,Yajima12} have all shown
systematic variations in the \ly line with the viewing angle. These
variations are a complex superposition of anisotropic density
configurations (i.e. edge-on vs. face-on view of a galaxy), the
inflows observed by gas cooling and the outflows included in the
supernova feedback process of the simulation. These bulk flows
physically correspond to the circumgalactic and intergalactic medium
(CGM and IGM). These effects are starting to be studied
 in simplified configurations that vary the density and wind
 characteristics \citep{Zheng2013,Behrens2014}.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.8\textwidth]{f1.eps}
\end{center}
\caption{Shape of the \ly line for
    different maximum rotational velocities. The continuous (dashed)
    line represents the central (homogeneous) source distributions. 
    These profiles were obtained integrating the flux of photons 
	over all directions, this means we made histograms of the 
	numbers of photons escaping the galaxy with a given frequency.
    \label{fig:differentvelocities}}  
\end{figure*}

@@

Are these features observable with current instrumental capabilities?
As an order of magnitude estimate we can consider that the width of
each one of the triple peaks is close to $1/3$ of the FWHM of the line
at a given $V_{\rm max}$. For the cases we have studied in this
subsection those values would be close to $50\kms$, $100\kms$ and
$200\kms$ for optical depths of $10^5$, $10^6$, and $10^7$
respectively. At a redshift of $z\sim 2-3$ each one of these widths
is multiplied by factors of $z+1\sim 3-4$. Considering a nominal value
for the resolution of $\Delta v=200\kms$ for a Near-IR spectrograph capable of observing
those galaxies \citep{Kulas12}, we find that only the case of triple
peaked emission for the most massive object $\tau_{\rm H}=10^{7}$,
$V_{\rm max}=200\kms$ would have a chance to be spectrally
resolved. We note that observations of triple peaked systems with
higher resolution spectrographs have been reported  by
\cite{Chonis2013}, however the large velocity off-sets ($\sim
1000$\kms) from the systemic redshift favor the hypothesis of an
interacting system, including emission from shocked and fluorecing
gas, as the origin of the triple peaked emission.

@@

\item In spite of the asymmetry induced by rotation the spatial
  anisotropy in the emission is usually below the
  $10\%$ level. This justifies our consideration of angle-averaged
  effects to quantify the emission line of rotating clouds.

@@
